Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-17 article.tex
rearrange code in parser level 0 writeup:  intersperse explanations between code chunks

Let's make a quick pass through a minimal parser combinator library so that we
can see them in action before we move on to the hard stuff.  Without any 
further ado, here's the code:
\begin{verbatim}
import Control.Applicative  (Applicative(..), Alternative(..))
import Control.Monad        (liftM, ap)

newtype Parser t a = 
    Parser {getParser :: [t] -> Maybe ([t], a)}

item :: Parser t t
item = Parser f
  where 
    f [] = Nothing
    f (x:xs) = Just (xs, x)

@@

instance Applicative (Parser t) where
  pure = return
  (<*>) = ap

instance Alternative (Parser t) where
  empty = Parser (const empty)
  Parser l <|> Parser r = Parser (\ts -> l ts <|> r ts)

-- similar to mfilter, but for Alternative instead of MonadPlus
check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
check p m =
    m >>= \x -> if p x then return x else empty

literal :: Eq t => t -> Parser t t
literal x = afilter (== x) item
\end{verbatim}
First, we define the Parser type, which says that a Parser is a function
of a list of tokens, producing a new list of tokens and a result
value, and possibly failing.
Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
consume a single token if available, and fail otherwise.
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
For choosing between alternatives, the Alternative type class will do, so we create
an instance that is defined partly in terms of the Alternative instance of the
underlying result type, \verb+Maybe+.
Last are a couple of convenience combinators for creating specific token parsers.

Now for some examples:  first, let's see \verb+item+ in action.  It fails on empty lists,
and succeeds consuming one token otherwise:
\begin{verbatim}
*Main> getParser item ""
Nothing

@@

*Main> getParser (literal 'a' <|> literal 'b') "cdeab"
Nothing
\end{verbatim}
In the rest of this article, we'll be reworking these basic parser combinators.
We'll be simplifying the implementations, extending the types of the results,
all while taking advantage of monad transformers to keep it modular.

>>>>

Let's make a quick pass through a minimal parser combinator library so that we
can see them in action before we move on to the hard stuff.  There are three
main pieces of the code:
\begin{itemize}
 \item a Parser datatype
 \item parser-specific combinators
 \item type class instances
\end{itemize}

\subsection{Parser datatype}
\begin{verbatim}
newtype Parser t a = 
    Parser {getParser :: [t] -> Maybe ([t], a)}
\end{verbatim}
This type definition says that a Parser is a function
of a list of tokens, producing a new list of tokens and a result
value, and possibly failing.

\subsection{Parser-specific combinators}
\begin{verbatim}
item :: Parser t t
item = Parser f
  where 
    f [] = Nothing
    f (x:xs) = Just (xs, x)

literal :: Eq t => t -> Parser t t
literal x = afilter (== x) item
\end{verbatim}
Next is the \verb+item+ parser, the simplest parser we'll deal with.  All it does is 
consume a single token if available, and fail otherwise.

\subsection{Type class instances}
For running parsers in sequence, we'll use the combinators from the Monad and
Applicative type classes.  There's a little bit of book-keeping involved in the
Monad instance -- it has to make sure that the correct token list is used by
each parser -- but once we have a Monad instance, the Applicative and Functor
instances can be handled by imports \verb+ap+ and \verb+liftM+ from Control.Monad.
\begin{verbatim}
import Control.Applicative  (Applicative(..), Alternative(..))
import Control.Monad        (liftM, ap)

@@

instance Applicative (Parser t) where
  pure = return
  (<*>) = ap 
\end{verbatim}

For choosing between alternatives, the Alternative type class will do, so we 
create an instance that is defined in terms of the Alternative instance of the
underlying result type, \verb+Maybe+.
\begin{verbatim}
instance Alternative (Parser t) where
  empty = Parser (const empty)
  Parser l <|> Parser r = Parser (\ts -> l ts <|> r ts)

-- similar to mfilter, but for Alternative instead of MonadPlus
check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
check p m =
    m >>= \x -> if p x then return x else empty
\end{verbatim}

\subsection{Examples}
Now for some examples:  first, let's see \verb+item+ in action.  It fails on empty lists,
and succeeds consuming one token otherwise:
\begin{verbatim}
*Main> getParser item ""
Nothing

@@

*Main> getParser (literal 'a' <|> literal 'b') "cdeab"
Nothing
\end{verbatim}
In the rest of this article, we'll be reworking these basic parser combinators,
by improving the implementations and extending the result types,
all while taking advantage of monad transformers to keep it modular.
